---
title: "Final Project on Data Science for Health Systems"
author: "Sevasti Dimopoulou"
date: "2024-06-06"
output: html_document
---

# Libraries 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(lubridate)
```

# Loading the data
```{r}
# Loading the data
maternal_mortality <- read.csv("main_data.csv")
```


# Data Cleaning
```{r}
# Inspecting the structure and summary statistics of the data
str(maternal_mortality)
summary(maternal_mortality)
head(maternal_mortality)

# Selecting relevant columns
data <- maternal_mortality %>% 
  select(iso_alpha_3_code, year_start, year_end, year_mid, obs_matdeaths, final_pm, env_total, env_mat)

# Removing rows with missing values in key columns
filtered_data <- data %>% 
  filter(!is.na(obs_matdeaths) & !is.na(final_pm))

# Filtering years to a reasonable range
filtered_data <- filtered_data %>%
  filter(year_mid >= 1980 & year_mid <= 2020)

# Checking for outliers in 'obs_matdeaths' and 'final_pm'
boxplot(filtered_data$obs_matdeaths, main = "Boxplot for Observed Maternal Deaths")
boxplot(filtered_data$final_pm, main = "Boxplot for Proportional Mortality")

# Handling outliers 
# Here, I am capping outliers at the 99th percentile for illustration purposes

# Defininhg a function to cap outliers
cap_outliers <- function(x) {
  q <- quantile(x, probs = c(0.01, 0.99), na.rm = TRUE)
  x <- ifelse(x < q[1], q[1], ifelse(x > q[2], q[2], x))
  return(x)
}

# Appling the function to 'obs_matdeaths' and 'final_pm'
filtered_data$obs_matdeaths <- cap_outliers(filtered_data$obs_matdeaths)
filtered_data$final_pm <- cap_outliers(filtered_data$final_pm)

# Summary statistics after handling outliers
summary(filtered_data)

# Ensuring consistency in data types
filtered_data <- filtered_data %>%
  mutate(
    iso_alpha_3_code = as.factor(iso_alpha_3_code),
    year_start = as.integer(year_start),
    year_end = as.integer(year_end),
    year_mid = as.integer(year_mid),
    obs_matdeaths = as.integer(obs_matdeaths),
    final_pm = as.numeric(final_pm),
    env_total = as.numeric(env_total),
    env_mat = as.numeric(env_mat)
  )

```
```{r}
# Final inspection of cleaned data
str(filtered_data)
head(filtered_data)
```

# EDA

## Distribution of Observed Maternal Deaths
```{r}
ggplot(filtered_data, aes(x = obs_matdeaths)) +
  geom_histogram(binwidth = 100, fill = "blue", color = "black") +
  labs(title = "Distribution of Observed Maternal Deaths", x = "Observed Maternal Deaths", y = "Frequency")

```
## Distribution of Proportional Mortality
```{r}
ggplot(filtered_data, aes(x = final_pm)) +
  geom_histogram(binwidth = 0.005, fill = "green", color = "black") +
  labs(title = "Distribution of Proportional Mortality", x = "Proportional Mortality", y = "Frequency")


```


## Distribution of Environmental Total

```{r}
ggplot(filtered_data, aes(x = env_total)) +
  geom_histogram(binwidth = 1000, fill = "red", color = "black") +
  labs(title = "Distribution of Environmental Total", x = "Environmental Total", y = "Frequency")

```



## Distribution of Environmental Maternal
```{r}
ggplot(filtered_data, aes(x = env_mat)) +
  geom_histogram(binwidth = 100, fill = "purple", color = "black") +
  labs(title = "Distribution of Environmental Maternal", x = "Environmental Maternal", y = "Frequency")

```



##Total Maternal Deaths Over Time

```{r}
yearly_summary <- filtered_data %>% 
  group_by(year_mid) %>% 
  summarize(total_deaths = sum(obs_matdeaths), 
            average_pm = mean(final_pm, na.rm = TRUE))

ggplot(yearly_summary, aes(x = year_mid, y = total_deaths)) +
  geom_line() +
  labs(title = "Total Maternal Deaths Over Time", x = "Year", y = "Total Maternal Deaths")


```


## Average Proportional Mortality Over Time
```{r}
ggplot(yearly_summary, aes(x = year_mid, y = average_pm)) +
  geom_line() +
  labs(title = "Average Proportional Mortality Over Time", x = "Year", y = "Average PM")


```
## Summary Table for Countries
```{r}
summary_table <- filtered_data %>% 
  group_by(iso_alpha_3_code) %>% 
  summarize(mean_pm = mean(final_pm, na.rm = TRUE), 
            total_deaths = sum(obs_matdeaths)) %>%
  arrange(desc(total_deaths))

print(summary_table)

```

# Regional Analysis: Total Maternal Deaths Over Time

```{r}
# Filtering to top 10 countries with the most observations
top_countries <- filtered_data %>%
  group_by(iso_alpha_3_code) %>%
  summarize(count = n()) %>%
  top_n(10, wt = count) %>%
  pull(iso_alpha_3_code)

filtered_top_countries <- filtered_data %>%
  filter(iso_alpha_3_code %in% top_countries)

# Boxplot of Maternal Deaths by Top Countries
ggplot(filtered_top_countries, aes(x = reorder(iso_alpha_3_code, obs_matdeaths, FUN = median), y = obs_matdeaths)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Boxplot of Maternal Deaths by Top 10 Countries", x = "Country", y = "Maternal Deaths") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

## South America

```{r}
# Creating  the country_yearly_summary dataframe
country_yearly_summary <- filtered_data %>%
  group_by(iso_alpha_3_code, year_mid) %>%
  summarise(total_deaths = sum(obs_matdeaths, na.rm = TRUE))

ggplot(country_yearly_summary %>% filter(iso_alpha_3_code %in% c("BRA", "CHL", "ECU")), 
       aes(x = year_mid, y = total_deaths, color = iso_alpha_3_code)) +
  geom_line() +
  labs(title = "Total Maternal Deaths Over Time (South America)", x = "Year", y = "Total Maternal Deaths") +
  scale_color_discrete(name = "Country")

```

```{r}
# ASIA 
# Plotting total maternal deaths by year for multiple countries
ggplot(country_yearly_summary %>% filter(iso_alpha_3_code %in% c("MYS", "KAZ", "KOR")), 
       aes(x = year_mid, y = total_deaths, color = iso_alpha_3_code)) +
  geom_line() +
  labs(title = "Total Maternal Deaths Over Time (Multiple Countries)", x = "Year", y = "Total Maternal Deaths") +
  scale_color_discrete(name = "Country")

```
```{r}
# Comparative Analysis

# Plotting total maternal deaths by year for multiple countries
ggplot(country_yearly_summary %>% filter(iso_alpha_3_code %in% c("AUS", "USA", "GBR")), 
       aes(x = year_mid, y = total_deaths, color = iso_alpha_3_code)) +
  geom_line() +
  labs(title = "Total Maternal Deaths Over Time (Multiple Countries)", x = "Year", y = "Total Maternal Deaths") +
  scale_color_discrete(name = "Country")
```

```{r}
# Linear regression to analyze trends in proportional mortality
model <- lm(final_pm ~ year_mid, data = filtered_data)
summary(model)

# Plotting regression line
ggplot(filtered_data, aes(x = year_mid, y = final_pm)) +
  geom_point() +
  geom_smooth(method = "lm", col = "red") +
  labs(title = "Trend in Proportional Mortality Over Time", x = "Year", y = "Proportional Mortality")

```
## summary table
```{r}
summary_table <- filtered_data %>% 
  group_by(iso_alpha_3_code) %>% 
  summarize(mean_pm = mean(final_pm, na.rm = TRUE), 
            total_deaths = sum(obs_matdeaths)) %>%
  arrange(desc(total_deaths))

print(summary_table)
```









